A parametricity-based formalization of semi-simplicial and semi-cubical sets

### 2023-11-26 paper/paper.tex
paper: major work on last two sections

\subsection{Groupoid properties of equality and basic type isomorphisms\label{sec:eqproperties}}
The construction takes benefit of various provable equalities over proofs of equality being definitional by the reflection rule. This includes in particular the groupoid properties of equality. Notably, uniqueness of identity proofs holds in extensional type theory, so that any type is automatically an $\U$. Also, we left implicit in table~\ref{tab:coh} the use of the isomorphism between $u = v$ and $\Sigma (p:u.\hd = v.\hd). (u.\tl = v.\tl)$ for $u$ and $v$ in a $\Sigma$-type. In the same table, we also left implicit the use of the isomorphism between $f = g$ and $\Pi a: A.\, f(a) = g(a)$ for $f$ and $g$ in $\Pi a: A.\, B$, where it should be recalled that the right-to-left map, that is functional extensionality, holds by default in extensional type theory. As a final remark, note, as a consequence of $\eta$-conversion for finite enumerated types, that the requirement of functional extensionality disappears when $\nu$ is finite. However, this is a conversion which Coq does not implement, so, in Coq, the alternative would be to replace $\Pi a: \nu.\, B$ by a ``flat'' iterated product $B(1) \times B(2) \times \ldots \times B(\nu)$.

\section{Future work}
In the cubical case, we expect the construction to eventually provide a model of (some version of) parametric type theory~\citep{nuyts17,cavallo19} by adding degeneracies, a hierarchy of universes (as sketched e.g. in a talk by Herbelin at the HoTT-UF workshop for the bridge case~(\citeyear{herbelin-hott-uf})), as well as reasoning modulo permutations~\citep{grandis03}.

By equipping the universe construction with a structure of equivalences, as suggested along the lines of Altenkirch and Kaposi~(\citeyear{altenkirch15}), we expect the construction to be able to serve as a basis for syntactic models of various versions of cubical type theory~\citep{bezem13,cohen16,angiuli21}, saving the detour via the fibered approach inherent to usual presheaf models. This would a priori preserve definitional properties which may be lost when detouring via presheaves. In particular, we conjecture being able to justify univalence holding definitionally. Our approach would also definitively ground cubical type theory in iterated parametricity.

>>>>

\subsection{Groupoid properties of equality and basic type isomorphisms\label{sec:eqproperties}}
The construction relies on the groupoid properties of equality in some places, which have to be made explicit in the formalization.

Notice that in table~\ref{fulltab:coh}, the use of the isomorphism between $u = v$ and $\Sigma (p:u.\hd = v.\hd). (u.\tl = v.\tl)$ for $u$ and $v$ in a $\Sigma$-type is left implicit. Also implicit is the use of the isomorphism between $f = g$ and $\Pi a: A.\, f(a) = g(a)$ for $f$ and $g$ in $\Pi a: A.\, B$, where it should be recalled that the right-to-left map, or functional extensionality, holds by default in extensional type theory.

As a final remark, note that as a consequence of $\eta$-conversion for finite enumerated types, the requirement of functional extensionality disappears when $\nu$ is finite. However, this is a conversion which Coq does not implement, and the alternative would be to replace $\Pi a: \nu.\, B$ by a ``flat'' iterated product $B(1) \times B(2) \times \ldots \times B(\nu)$.

\section{Future work}
The construction could be extended with degeneracies as well as with permutations~\citep{grandis03}. Dependent $\nu$-sets could also be defined, opening the way to construct $\Pi$-types and $\Sigma$-types of $\nu$-sets. A $\nu$-set of $\nu$-set representing a universe could also be defined (as sketched e.g. in a talk by Herbelin at the HoTT-UF workshop for the bridge case~(\citeyear{herbelin-hott-uf})). More generally, we believe these lines of work to eventually provide alternative models to parametric type theories~\citep{nuyts17,cavallo19}.

By equipping the universe construction with a structure of equivalences, as suggested along the lines of Altenkirch and Kaposi~(\citeyear{altenkirch15}), we also suspect the construction to be able to serve as a basis for syntactic models of various versions of cubical type theory~\citep{bezem13,cohen16,angiuli21}, saving the detour via the fibered approach inherent to usual presheaf models. In particular, we wonder whether the indexed approach might provide definitional properties which are otherwise lost when detouring via presheaves. In particular, we conjecture being able to justify univalence holding definitionally. Our approach would also definitively ground cubical type theory in iterated parametricity.
